\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2020-12-09 paper.tex
lights v2

\usepackage{mathtools}
\usepackage{xcolor}
\usepackage{mathrsfs} 
\usepackage{natbib}
\usepackage{setspace}
\usepackage{dsfont}
\usepackage{caption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{booktabs}

@@

@@

\newcommand{\dd}{\mathrm{d}}
\newcommand{\ind}[1]{\mathds{1}_{#1}}
\newcommand{\norm}[1]{\|#1\|}
\newcommand{\cM}{\mathcal M}
\newcommand{\cD}{\mathcal D}
\newcommand{\cC}{\mathcal C}
\newcommand{\cN}{\mathcal N}
\newcommand{\cA}{\mathcal A}
\newcommand{\R}{\mathds R}
\newcommand{\N}{\mathds N}
\newcommand{\E}{\mathds E}
\renewcommand{\P}{\mathds P}
\newcommand{\bSigma}{\textbf{$\Sigma$}}

@@

This paper introduces a prognostic method called \textit{lights} to deal with the problem of joint modeling of longitudinal data and censored durations, where a large number of longitudinal features are available. Yet there is no standard model so far to learn from such high-dimensional multivariate longitudinal data in a survival analysis setting.
The statistical performance of the method is examined on an extensive Monte Carlo simulation study, and finally illustrated on a publicly available dataset.
On these high-dimensional datasets, our proposed method significantly outperforms the state-of-the-art survival models regarding risk prediction in terms of C-index, with a computing time orders of magnitude faster. In addition, it provides powerful interpretability by automatically pinpointing significant features being relevant from a clinical perspective. Thus, we propose a powerfull tool for personalized medicine, with the ability of automatically determining significant prognostic longitudinal biomarkers, which is of increasing importance in many areas of medicine.
\\

@@

In many clinical studies, it has become increasingly common to record the values of longitudinal features (e.g. biomarkers) until the occurrence of an event of interest for a subject. The ``joint modeling'' approaches, namely modeling the longitudinal and survival outcomes through a joint likelihood model rather than separately, has received considerable attention during the past two decades~\citep{wulfsohn1997joint}. Numerical studies suggest that these approaches are among the most satisfactory to combine information~\citep{tsiatis2004joint,yu2004joint}. They have the advantage of making more efficient use of the data since information about survival also goes into modeling the longitudinal features. In addition, they produce unbiased estimates and do not rely on approximations for incorporating conplex longitudinal trajectories.

\paragraph{The high-dimensional longitudinal data context.}
With the development of electronic health records, high-dimensional settings are becoming increasingly frequent in various contexts where the number of available features to consider as potential risk factors is tremendous.
Moreover, with an increased focus on personalised medicine, the need to implement multivariate models that account for a large number of longitudinal outcomes is critical. Despite this, joint models have predominantly focused on univariate data, with attempts to fit multiple univariate joint models separately~\citep{wang2012joint}, which is inefficient~\citep{lin2002maximum}. 
Despite many multivariate models being presented in full generality, questions arrising from the high-dimensional context -- computational power, limits in numerical estimation, or sample size for instance -- are never considered in analyses (to the best of our knowledge), and the number of longitudinal outcomes considered in numerical studies are often very low (see~\citet{hickey2016joint} for a complete review). For instance,~\citet{jaffa2014joint} only considers 3 longitudinal outcomes in the simulation study while mentioning a ``high-dimensional multivariate longitudinal data'' context.

@@

The setting of this paper is such that we want to incorporate high-dimensional time-dependent (longitudinal) features measured with error in a survival model. Let us consider the usual survival analysis framework.
Following~\citet{AndBorGilKei-93}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively. The event of interest could be for instance survival time, re-hospitalization, relapse or disease progression.
We then denote $T$ the right-censored time and $\Delta$ the censoring indicator, defined as 
\begin{equation*}
T = T^\star \wedge C \quad \text{and} \quad \Delta = \ind{}({\{T^\star \leq C\}})
\end{equation*}
respectively, where $a \wedge b$ denotes the minimum between two numbers $a$ and $b$, and $\ind{}(\cdot)$ the indicator function taking the value $1$ if the condition in $(\cdot)$ is satisfied and $0$ otherwise.

Let $X$ denotes the $p$-dimensional vector of time-independent features (e.g. patients characteristics, therapeutic strategy, or omics features recorded at the begining of a study), and let  $Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L$ denote the value of the $L$-dimensional longitudinal outcome at time point $t \geq 0$, with $L > 0$.

@@

An assumption of heterogeneity within the patient population is frequently relevant in medical research where several differing profiles of subjects are expected~\citep{bussy2019c}. To take account of this, we introduce a latent variable $G \in \{0, \ldots, K-1\}$ modeling the $K \geq 1$ subgroups of different risk. Let us denote
\[ \pi_{\xi_k}(x) = \P[G=k|X=x] \]
the latent class membership probability given time-independent features $x \in \R^p$, and consider a softmax link function given by
\[ \pi_{\xi_k}(x) = \dfrac{e^{x^\top\xi_k}}{\sum_{k=0}^{K-1}e^{x^\top\xi_k}} \]
where $\xi_k \in \R^p$ denotes a vector of coefficients that quantifies the impact of each time-independent features on the probability that a subject belongs to the $k$-th group.  From now on, all computations are done conditionally on features $x$.

Then, the idea is to allow flexibility in modeling the dependency between the longitudinal features and the event time.
Joint models have the ability to produce prognostic predictions that can be dynamically updated according to the observed trajectory of the features, thus offering a powerful tool for clinical decision making in patient monitoring. But those questions are never considered in a high-dimensional framework~\citep{proust2009development,rizopoulos2011dynamic}.

When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival}, or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
Developments in the field of joint modeling have primarily focused on modeling and estimation, and most studies do not consider goodness of fit nor generalisation power [REF!]. However, practitioners will naturally require predictive prognostic tools to evaluate model fits and compare models.

Once the model is train, posterior risk classification can be made through posterior estimates of the latent class membership probabilities, and latent class model may be particularly suited for prediction problems. We propose to use this probabilities as discriminative marker rule for risk prediction and use it in the cross-validation procedure for selecting the best regularization hyper-parameters. Indeed, it makes sense from a practical point of view, since one want to obtain prognostic predictions as well as interpretations of the high-dimensional longitudinal features.

@@

In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) that \texttt{blabla... ajout description courte des caractéristiques du modèle une fois les maths terminées}.

@@

A precise description of the model is given in Section~\ref{sec:Method}.
Section~\ref{sec:inference} focuses on a regularized version of the
model to exploit dimension reduction and prevent overfitting. Inference is
presented under this framework, as well as the convergence properties of the developed algorithm.
Section~\ref{sec:Performance evaluation} presents the simulation procedure used to evaluate the performance of our method and compares it with state-of-the-art ones. In Section~\ref{sec:application}, we apply our method to high-dimensional publicly available datasets. Finally, we discuss the obtained results in Section~\ref{sec:conclusion}.

@@

Throughout the paper, for every $q > 0,$ we denote by $\norm{v}_q$ the usual $\ell_q$-quasi norm of a vector $v \in \R^m,$ namely
 $\norm{v}_q =(\sum_{k=1}^m|v_k|^q)^{1/q}$. We also denote $\norm{v}_0 = |\{k : v_k \neq 0\}|$, where $|A|$ stands for the cardinality of a finite set $A$.
 For $u, v \in \R^m$, we denote by $u \odot v$ the Hadamard product $u\odot v =(u_1v_1, \ldots, u_mv_m)^\top.$ 
For a squared matrix $M$, $\text{vech}(M)$ stacks columns of $M$ one under another in a single vector, starting each column at its diagonal element.
Finally, we write, for short, $\mathbf{1}_m$  (resp. $\mathbf{0}_m$) for the vector of $\R^m$ having all coordinates equal to one (resp. zero).

@@

\paragraph{Group-specific marker trajectories.}

We suppose a group-specific marker trajectory and a generalized linear mixed model for each longitudinal marker given subgroup $G$, so that for the $l$-th outcome at time $t \geq 0$ one has
\begin{equation}
	\label{eq:link-function}
 	h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t)
 \end{equation}
where $h_l$ denotes a known one-to-one link function, and $m_k^l$ the linear predictor such that
\[ m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l \]
where $u^l(t) \in \R^{p_l}$ is a row vector of (possibly) time-varying features with corresponding unknown fixed effect parameters $\beta_k^l$, and $v^l(t) \in \R^{r_l}$ is a row vector of (possibly) time-varying features with $r_l \leq p_l$ and where the corresponding subject-and-longitudinal outcome specific random effects $b^l$.
A suitable distributional assumption for the random effects component is a zero-mean multivariate normal distribution~\citep{hickey2016joint}, that is
\[ b^l \sim \cN(0, D_{ll}) \]
with $D_{ll} \in \R^{r_l \times r_l}$ the unstructured variance-covariance matrix that does not depend on the group membership. 
To account for dependence between the different longitudinal outcome types, we let $\text{Cov}[b^l,b^{l'}] = D_{ll'}$ for $l \ne l'$ and we denote
\[ D = 
\begin{bmatrix}
  D_{11} & \cdots & D_{1L}\\
  \vdots &  \ddots & \vdots \\
  D_{1L}^\top & \cdots & D_{LL}
\end{bmatrix}
\]
the global variance-covariance matrix.
In the sequel of the paper, our aim is to associate the true and unobserved value $m_k^l(t)$ of the $l$-th longitudinal outcome at time $t$ with the event outcome $T^\star$.

\paragraph{Group-specific risk of event.}

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a relative risk model of the form
\begin{equation}
	\label{eq:intensity-model}
	\lambda(t|\cM_k(t), G = k, w) = \lambda_0(t) \exp \Big\{ w^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA \gamma_{k,a}^l \varphi_a(t, \beta_k^l, b^l) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ denotes an unspecified baseline hazard function and given that $G = k$, $\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}$ denotes the history of the true unobserved longitudinal process up to time $t$, $w$ is a $q$-vector of time-independant features with corresponding fixed effects $\gamma_k \in \R^q$, and $\gamma_{k,a}^l \in \R$ is the joint model association parameter corresponding to a known functional $\varphi_a$ defining a shared association. Details on the choice of $(\varphi_a)_{a \in \cA}$ are given in Section~\ref{sec:numerical details}. From now on, all computations are done conditionally on features $w$.

\paragraph{Likelihood.}

Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
\[ \cD = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
\[ y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) \]
for all $l=1, \ldots, L$. 
For the $i$-th subject, let us denote 
\[
\left\{
    \begin{array}{ll}
        y_i &= ({y_i^1}^\top, \ldots, {y_i^L}^\top)^\top \in \R^{n_i},\\
        b_i &= ({b_i^1}^\top, \ldots, {b_i^L}^\top)^\top \in \R^r,
    \end{array}
\right.
\]
with $n_i = \sum_{l=1}^L n_i^l$ and $r = \sum_{l=1}^L r_l$ the total number of longitudinal measurements (for subject $i$) and the total dimension of the random effects respectively, as well as the following design matrices
\[ U_i = 
\begin{bmatrix}
  U_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & U_{iL}
\end{bmatrix} 
\in \R^{n_i \times p}
\qquad \text{and} \qquad
V_i = 
\begin{bmatrix}
  V_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & V_{iL}
\end{bmatrix}
\in \R^{n_i \times r}
\]
where for all $l=1, \ldots, L$, one writes
\[
\left\{
    \begin{array}{ll}
        U_{il} &= [u_i^l(t_{i1}^l) \cdots u_i^l(t_{in_i^l}^l)]^\top \in \R^{n_i^l \times p_l},\\
        V_{il} &= [v_i^l(t_{i1}^l) \cdots v_i^l(t_{in_i^l}^l)]^\top \in \R^{n_i^l \times r_l}.
    \end{array}
\right.
\]
Note that from now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
Suppose that conditional on the random effects, each $Y_i^l$ is independent and the censoring times $T_i$ are independent of $b_i$; both are standard modelling assumptions.
For all $k = 0, \ldots, K-1$, we denote 
\[ \beta_k = ({\beta_k^1}^\top, \ldots, {\beta_k^L}^\top) \in \R^p \] 
with $p = \sum_{l=1}^L p_l$ and 
\[M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}. \]

Given~\eqref{eq:link-function}, each $y_i^l$ is assumed to be from a one-parameter exponential family with respect to a reference measure which is either the Lebesgue measure (e.g. in the Gaussian case) or the counting measure (e.g. in the logistic cases).
The conditional distribution of $y_i|b_i, G_i=k$ is then assumed to be from a distribution with a density of the form
\begin{equation}
	\label{eq:exp-family-density}
	f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\},
\end{equation}
with 
\[ \Phi_i = [\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top]^\top \in \R^{n_i} \] 
and $\phi = (\phi_1, \ldots, \phi_L)^\top \in \R^L$.
The density described in~\eqref{eq:exp-family-density} encompasses several distributions, see Table~\ref{table:glm}. The functions $c_\phi(\cdot)$ and $d_\phi(\cdot)$ are known as well as the dispersion parameters $\phi_l$, while parameters $\beta_k$ have to be estimated. Note that $d_\phi(\cdot)$ is related to the normalizing constant.
\begin{table}[htb]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{cccccc}
\toprule
 Model & Support & Use cases & $\phi_l$ & $h_l(\cdot)$ in~\eqref{eq:link-function} & $c_{\phi_l}(\cdot)$ \\
 \midrule
 Normal & $\R$ & Continuous response data & $\sigma_l^2$ & $z \mapsto z$ & $z \mapsto \phi_l^{-1}\frac{z^2}{2}$ \\ 
 Categorical & $\{0, 1\}^C$ & Outcome with $C$ modalities & 1 & $z \mapsto \log\Big(\dfrac{z}{1 - z} \Big)$ & $z \mapsto \phi_l^{-1} \frac{e^z}{1+e^z}$ \\
 Poisson & $\N$ & Count of occurrences & 1 &  $z \mapsto e^z$ & $z \mapsto \phi_l^{-1} e^z$ \\
 \bottomrule
\end{tabular}}
\caption[table]{\small Examples of standard distributions that fit in the considered setting.}
\label{table:glm}
\end{table}

>>>>

@@

\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{xcolor}
\usepackage{mathrsfs} 
\usepackage{natbib}
\usepackage{setspace}
\usepackage{dsfont}
\usepackage{caption}
\usepackage{booktabs}

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\newcommand\mycommfont[1]{\footnotesize\ttfamily\textcolor{orange}{#1}}
\SetCommentSty{mycommfont}
\SetKwInput{KwInput}{Input} 
\SetKwInput{KwOutput}{Output}

\newtheorem{assumption}{Assumption}{\bf}{\rm}

@@

\newcommand{\dd}{\mathrm{d}}
\newcommand{\ind}[1]{\mathds{1}_{#1}}
\newcommand{\norm}[1]{\|#1\|}
\newcommand{\cM}{\mathcal M}
\newcommand{\cD}{\mathcal D}
\newcommand{\cC}{\mathcal C}
\newcommand{\cN}{\mathcal N}
\newcommand{\cA}{\mathcal A}
\newcommand{\cQ}{\mathcal Q}
\newcommand{\R}{\mathds R}
\newcommand{\N}{\mathds N}
\newcommand{\E}{\mathds E}
\renewcommand{\P}{\mathds P}
\newcommand{\bSigma}{\textbf{$\Sigma$}}

@@

This paper introduces a prognostic method called \textit{lights} to deal with the problem of joint modeling of longitudinal data and censored durations, where a large number of longitudinal features are available. Yet there is no standard model so far to learn from such high-dimensional multivariate longitudinal data in a survival analysis setting.
Features are extracted from the longitudinal processes and included as potential risk factor in a group-specific Cox model with high-dimensional shared associations. Appropriate penalties are then used during inference to allow flexibility in modeling the dependency between the longitudinal features and the event time.
The statistical performance of the method is examined on an extensive Monte Carlo simulation study, and finally illustrated on publicly available datasets.
On these high-dimensional datasets, our proposed method significantly outperforms the state-of-the-art survival models regarding risk prediction in terms of C-index, with a computing time orders of magnitude faster. In addition, it provides powerful interpretability by automatically pinpointing significant features being relevant from a clinical perspective. Thus, we propose a powerfull tool for personalized medicine, with the ability of automatically determining significant prognostic longitudinal biomarkers, which is of increasing importance in many areas of medicine.
\\

@@

In many clinical studies, it has become increasingly common to record the values of longitudinal features (e.g. biomarkers) until the occurrence of an event of interest for a subject. The ``joint modeling'' approaches, namely modeling the longitudinal and survival outcomes through a joint likelihood model rather than separately, has received considerable attention during the past two decades~\citep{tsiatis2004joint}. Numerical studies suggest that these approaches are among the most satisfactory to combine information~\citep{yu2004joint}. They have the advantage of making more efficient use of the data since information about survival also goes into modeling the longitudinal features. In addition, they produce unbiased estimates and do not rely on approximations for incorporating conplex longitudinal trajectories. Most developments have either focused on shared random-effect models (SREMs)~\citep{wulfsohn1997joint}, in which characteristics of the longitudinal processes (for instance functions of the random effects) are included as features in the survival model ; or on joint latent class models (JLCMs)~\citep{vermunt2003latent}, in which the population is considered as heterogeneous, with the assumption that there exist homogeneous latent subgroups sharing the same marker trajectories and the same prognostic.

\paragraph{The high-dimensional longitudinal data context.}
With the development of electronic health records, high-dimensional settings are becoming increasingly frequent in various contexts where the number of available features to consider as potential risk factors is tremendous.
Moreover, with an increased focus on personalised medicine, the need to implement multivariate models that account for a large number of longitudinal outcomes is critical. Despite this, joint models have predominantly focused on univariate data, with attempts to fit multiple univariate joint models separately~\citep{wang2012joint}, which is inefficient~\citep{lin2002maximum}. 
Despite many multivariate models being presented in full generality, questions arrising from the high-dimensional context -- e.g. computational power, limits in numerical estimation, or sample size -- are never considered in analyses (to the best of our knowledge), and the number of longitudinal outcomes considered in numerical studies are often very low (see~\citet{hickey2016joint} for a complete review). For instance,~\citet{jaffa2014joint} only considers 3 longitudinal outcomes in the simulation study while mentioning a ``high-dimensional multivariate longitudinal data'' context.

@@

The setting of this paper is such that we want to incorporate high-dimensional time-dependent (longitudinal) features measured with error in a survival model. Let us consider the usual survival analysis framework.
Following~\citet{AndBorGilKei-93}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively. The event of interest could be for instance survival time, re-hospitalization, relapse or disease progression.
We then denote $T$ the right-censored time and $\Delta$ the censoring indicator, defined as 
\begin{equation*}
T = T^\star \wedge C \quad \text{and} \quad \Delta = \ind{\{T^\star \leq C\}}
\end{equation*}
respectively, where $a \wedge b$ denotes the minimum between two numbers $a$ and $b$, and $\ind{\{\cdot\}}$ the indicator function taking the value $1$ if the condition in $\{\cdot\}$ is satisfied and $0$ otherwise.

Let $X$ denotes the $p$-dimensional vector of time-independent features (e.g. patients characteristics, therapeutic strategy, or omics features recorded at the begining of a study), and let  $Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L$ denote the value of the $L$-dimensional longitudinal outcome at time point $t \geq 0$, with $L \in \N_+$.

@@

An assumption of heterogeneity within the patient population is frequently relevant in medical research where several differing profiles of subjects are expected~\citep{bussy2019c}. To take account of this, we introduce a latent variable $G \in \{0, \ldots, K-1\}$ modeling the $K \geq 1$ subgroups of different risk, which is a classical modeling assumption in JLCMs~\citep{lin2002latent, proust2014joint}. Let us denote
\begin{equation}
  \label{eq:pi}
  \pi_{\xi_k}(x) = \P[G=k|X=x]
\end{equation}
the latent class membership probability given time-independent features $x \in \R^p$, and consider a softmax link function given by
\[ \pi_{\xi_k}(x) = \dfrac{e^{x^\top\xi_k}}{\sum_{k=0}^{K-1}e^{x^\top\xi_k}} \]
where $\xi_k \in \R^p$ denotes a vector of coefficients that quantifies the impact of each time-independent features on the probability that a subject belongs to the $k$-th group, with $\xi_0 = \mathbf{0}_p$ for overparameterization purpose, where $\mathbf{0}_p$ stands for the vector of $\R^p$ having all coordinates equal to zero. The intercept term is here omitted without loss of generality. 
From now on, all computations are done conditionally on features $x$.

@@

In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) which is from both JLCMs and SREMs, since we also include features extracted from the longitudinal processes as potential risk factor in the survival model, which is a group-specific Cox model~\citep{Cox1972JRSS} with high-dimensional shared associations.
To allow flexibility in modeling the dependency between the longitudinal features and the event time, we use appropriate penalties : elastic net~\citep{zou2005regularization}
for feature selection in the latent class membership as well as within the survival model, and sparse group lasso~\citep{simon2013sparse} for the fixed effect of each marker trajectories, allowing flexible representations of time.
Inference is achieved using an efficient Quasi-Newton Expectation Maximization algorithm.
The posterior estimates of the latent class membership probabilities are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters to adapt to the data complexity.
Hence, the method provides interpretations of the high-dimensional longitudinal features, thus offering a powerful tool for clinical decision making in patient monitoring.

@@

A precise description of the model is given in Section~\ref{sec:Method}.
Section~\ref{sec:inference} focuses on a regularized version of the
model to exploit dimension reduction and prevent overfitting. Inference is
presented under this framework, as well as the developed algorithm.
Section~\ref{sec:Performance evaluation} introduces the C-index metric, as well as a novel evaluation strategy to assess diagnostic prediction performances while mimicking a real-time use of the model in clinical care, and finally the competing methods considered.
Section~\ref{sec:simulation study} presents the simulation procedure used to evaluate the performance of our method in a high-dimensional context and compares it with state-of-the-art ones. In Section~\ref{sec:application}, we apply our method to high-dimensional publicly available datasets. Finally, we discuss the obtained results in Section~\ref{sec:conclusion}.

@@

Throughout the paper, for every $q > 0,$ we denote by $\norm{v}_q$ the usual $\ell_q$-quasi norm of a vector $v \in \R^m,$ namely
 $\norm{v}_q =(\sum_{k=1}^m|v_k|^q)^{1/q}$. We also denote $\norm{v}_0 = |\{k : v_k \neq 0\}|$, where $|A|$ stands for the cardinality of a finite set $A$.
 For $u, v \in \R^m$, we denote by $u \odot v$ the Hadamard product $u\odot v =(u_1v_1, \ldots, u_mv_m)^\top.$ 
For a squared matrix $M$, $\text{vech}(M)$ stacks columns of $M$ one under another in a single vector, starting each column at its diagonal element. We write $I_m$ for the identity matrix of $\R^{m \times m}$.
Finally, we write, for short, $\mathbf{1}_m$  (resp. $\mathbf{0}_m$) for the vector of $\R^m$ having all coordinates equal to one (resp. zero).

@@

\subsection{Group-specific marker trajectories}

We suppose a group-specific marker trajectory and a generalized linear mixed model for each longitudinal marker given subgroup $G$, so that for the $l$-th outcome at time $t \geq 0$ one has
\begin{equation}
	\label{eq:link-function}
 	h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t)
 \end{equation}
where $h_l$ denotes a known one-to-one link function, and $m_k^l$ the linear predictor such that
\[ m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l \]
where $u^l(t) \in \R^{q_l}$ is a row vector of (possibly) time-varying features with corresponding unknown fixed effect parameters $\beta_k^l$, and $v^l(t) \in \R^{r_l}$ is a row vector of (possibly) time-varying features with $r_l \leq q_l$ and where the corresponding subject-and-longitudinal outcome specific random effects $b^l$ that does not depend on the group membership, which is not a strong modeling assumption.
\begin{assumption}
\label{assumption1}
We suppose that the random effects are independent of the group membership, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
\end{assumption}
A suitable distributional assumption for the random effects component is a zero-mean multivariate normal distribution~\citep{hickey2016joint}, that is
\[ b^l \sim \cN(0, D_{ll}) \]
with $D_{ll} \in \R^{r_l \times r_l}$ the unstructured variance-covariance matrix. 
To account for dependence between the different longitudinal outcome types, we let $\text{Cov}[b^l,b^{l'}] = D_{ll'}$ for $l \ne l'$ and we denote
\[ D = 
\begin{bmatrix}
  D_{11} & \cdots & D_{1L}\\
  \vdots &  \ddots & \vdots \\
  D_{1L}^\top & \cdots & D_{LL}
\end{bmatrix}
\]
the global variance-covariance matrix.
In the sequel of the paper, our aim is to associate the true and unobserved value $m_k^l(t)$ of the $l$-th longitudinal outcome at time $t$ with the event outcome $T^\star$.

\subsection{Group-specific risk of event}

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
	\label{eq:intensity-model}
	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{ x^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, \beta_k^l, b^l) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
\[\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}\] 
the history of the true unobserved longitudinal process up to time $t$, $x$ is the $p$-vector of time-independant features (also used in~\eqref{eq:pi}, with no \textit{a priori} on the choice of features involved in the definition of $\pi_{\xi_k}$ nor $\lambda$, since independant regularizations are used during the inference, see~\eqref{eq:pen-log-lik}) with corresponding fixed effects $\gamma_{k,0} \in \R^p$, and for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\rho_a}$ the corresponding joint association parameters, and $\rho_a \in \N_+$ the dimension of the corresponding $\varphi_a$. 
This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
\[\gamma_k= (\gamma_{k,0}^\top, \gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{p+L\cA}. \]

The association structure between the longitudinal and the time-to-event submodels is key to the joint modeling framework. In spite of that, rationale for selecting shared associations has received little attention. We then propose to include some of the most common parameterization with no \textit{a priori}, set out in Table~\ref{table:shared_associations}, and let the model select the relevant ones through the regularization strategy described in Section~\ref{sec:penalized-obj}.
\begin{table}[htb]
\centering
\begin{tabular}{cccc}
\toprule
Description & $\varphi_a(t, \beta_k^l, b^l)$ & $\rho_a$ & Reference \\
\midrule
Linear predictor & $m_k^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
Random effects & $b^l$ & $r_l$ & \citet{hatfield2011joint} \\ 
Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
 \bottomrule
\end{tabular}
\caption[]{Description of the shared associations we include in the group-specific risk of event submodel~\eqref{eq:intensity-model}.}
\label{table:shared_associations}
\end{table}

\subsection{Likelihood}

Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
\[ y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) \]
for all $l=1, \ldots, L$. 
For the $i$-th subject, let us denote 
\[
\left\{
    \begin{array}{ll}
        y_i &= ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i},\\
        b_i &= ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r,
    \end{array}
\right.
\]
with $n_i = \sum_{l=1}^L n_i^l$ and $r = \sum_{l=1}^L r_l$ the total number of longitudinal measurements (for subject $i$) and the total dimension of the random effects respectively, as well as the following design matrices
\[ U_i = 
\begin{bmatrix}
  U_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & U_{iL}
\end{bmatrix} 
\in \R^{n_i \times q}
\qquad \text{and} \qquad
V_i = 
\begin{bmatrix}
  V_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & V_{iL}
\end{bmatrix}
\in \R^{n_i \times r}
\]
with $q = \sum_{l=1}^L q_l$ and where for all $l=1, \ldots, L$, one writes
\[
\left\{
    \begin{array}{ll}
        U_{il} &= \big(u_i^l(t_{i1}^l)^\top \cdots u_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times q_l},\\
        V_{il} &= \big(v_i^l(t_{i1}^l)^\top \cdots v_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times r_l}.
    \end{array}
\right.
\]
From now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
\begin{assumption}
Suppose that conditional on the random effects, each $Y_i^l$ is independent, and that the censoring times $T_i$ are independent of $b_i$.
\end{assumption}
This is a standard modeling assumption~\citep{tsiatis2004joint}.
For all $k = 0, \ldots, K-1$, we denote 
\[ \beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q \] 
and 
\[M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}. \]
